---
title: "Forecast data integration with observed"
author: "ROSSyndicate"
date: "2024-12-01"
output: html_document
---

# Purpose

This markdown document is intended to determine how to handle the forecast
data obtained from NOAA GEFS using Herbie to integrate as part of the data
assimilation pipeline for operational forecasting of temperature in SMR.

## Load observed data and forecasted data

First, the observed met data - this is loaded from the NASA-NW repository
workflow.

```{r}
library(tidyverse)
library(feather)

w_SMR_today_met <- read_feather("~/Desktop/w_SMR_today_met.feather")
g_SMR_met_QAQC <- read_feather("~/Desktop/g_SMR_met_QAQC.feather")

```

For our forecasts, we need to read in all the files, collate them, and then 
just grab the 'right now' and the 6h to compare sw/lw radiation.

```{r}
all_forecast_files <- list.files("data/herbie_extraction/debias/", full.names = T)

if (file.exists("data/herbie_extraction/compiled_forecast/GEFS_0p25_control_raw_2020-2024.feather")) {
  control <- read_feather("data/herbie_extraction/compiled_forecast/GEFS_0p25_control_raw_2020-2024.feather")
} else {
  control <- map(all_forecast_files, read_csv) %>% bind_rows()

  write_feather(control, "data/herbie_extraction/compiled_forecast/GEFS_0p25_control_raw_2020-2024.feather")
}

```

We can drop a few columns here that we don't really care about, and grab 00 step 
for all vars except SR, which we have to grab from the 03 (no 00 data).

```{r}
forecast_control <- control %>% 
  filter(number == 0, str_sub(step, 8, 9) == '00') %>% 
  select(time, t2m, tmin, tmax, u10, v10, r2) %>% 
  # convert to local time
  mutate(time = time - hours(7)) %>% 
  slice(1, .by = time)
forecast_control_sr <- control %>% 
  filter(number == 0, str_sub(step, 8, 9) == '03') %>% 
  select(time = valid_time, sdswrf) %>% 
  # convert to local time
  mutate(time = time - hours(7)) %>% 
  slice(1, .by = time)
```

Now, to SI:
- t2m is in deg K, needs to go to C
- u10,  v10 need to be transformed into wind speed

```{r}
forecast_control <- forecast_control %>% 
  mutate(fore_temp_degC = (t2m - 273.15),
         fore_wind_mps = sqrt(u10^2 + v10^2), #sqrt(mer^2 + zon^2)
         fore_rh = r2) %>%  
  select(-c(t2m, u10, v10, r2, tmin, tmax)) 
forecast_conrol_sr <- forecast_control_sr %>% 
  rename(fore_sr = sdswrf)
```

## Look for biases data

```{r}
prev_3hour <- g_SMR_met_QAQC %>% 
  mutate(date = date(datetime),
         hour = hour(datetime),
         # need to label each 3-hour segment by the hour we want to assign
         # the summary to at 2, 8, 14, 20
         hour3 = case_when(between(hour, 0, 1) | hour == 23 ~ 2,
                           between(hour, 5, 7) ~ 8,
                           between(hour, 11, 13) ~ 14,
                           between(hour, 17, 19) ~ 20),
         # shift 23h to next day for summary
         date = if_else(hour == 23, date + days(1), date)) %>% 
  filter(parameter  == "Solar_Radiation_Total", !is.na(hour3)) %>% 
  pivot_wider(names_from = parameter,
              values_from = value) %>% 
  mutate(time = ymd_h(paste(date, hour3, sep = " "))) %>% 
  summarize(obs_sr = mean(Solar_Radiation_Total),
            .by = time) 
# we need 3h intervals of data to match our forecast data
# instantaneous data: rh, wind, temp
instantaneous <- g_SMR_met_QAQC %>% 
  filter(hour(datetime) %in% seq(5, 23, 6)) %>% 
  filter(parameter %in% c("Air_Temperature_Avg", 
                          "Wind_Speed_Avg", 
                          "Relative_Humidity_Avg")) %>% 
  pivot_wider(names_from = parameter,
              values_from = value) %>% 
  rename(obs_air_temp = Air_Temperature_Avg,
         obs_wind = Wind_Speed_Avg,
         obs_rh = Relative_Humidity_Avg) %>% 
  rename(time = datetime)
  
```

collate the control forecast with observed 

```{r}
bias_comp_inst <- left_join(forecast_control, instantaneous)
bias_comp_sr <- left_join(forecast_conrol_sr, prev_3hour)
```


### Air temp

```{r}
ggplot(bias_comp_inst, aes(x = obs_air_temp, y = fore_temp_degC)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0)

ggplot(bias_comp_inst, aes(x = obs_air_temp, y = fore_temp_degC)) +
  geom_point() +
  facet_grid(hour(time) ~ .)+
  geom_abline(slope = 1, intercept = 0)

```

These look good, though generally the 17h temp is warmer than the forecast.

### Wind

```{r}
ggplot(bias_comp_inst, aes(x = obs_wind, y = fore_wind_mps)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0)

ggplot(bias_comp_inst, aes(x = obs_wind, y = fore_wind_mps)) +
  geom_point() +
  facet_grid(hour(time) ~ .)+
  geom_abline(slope = 1, intercept = 0)

```

Meh, not great, not terrible.

### Relative humidity

```{r}
ggplot(bias_comp_inst, aes(x = obs_rh, y = fore_rh)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0)

ggplot(bias_comp_inst, aes(x = obs_rh, y = fore_rh)) +
  geom_point() +
  facet_grid(hour(time) ~ .)+
  geom_abline(slope = 1, intercept = 0)

```

These look good, other than slightly higher rh in the AM hours.

### Solar radiation

And now we'll look at the control at 3h for solar radiation. This isn't perfect,
but it gives us an idea of approximate bias

```{r}
ggplot(bias_comp_sr, aes(x = obs_sr, y = fore_sr)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0)

ggplot(bias_comp_sr, aes(x = obs_sr, y = fore_sr)) +
  geom_point() +
  facet_grid(hour(time) ~ .)+
  geom_abline(slope = 1, intercept = 0)

```

These actually look pretty good too. Hopefully the 8 and 20 hour will offset each 
other. So these are g2g.


