---
title: "01-Collate Forecast Data"
author: "ROSSyndicate"
date: "2024-12-01"
output: html_document
---

```{r}
library(tidyverse)
library(feather)
```

# Purpose

This markdown document collates 3h GEFS 0.25 data from our Herbie download for 
all 3h data.

For our forecasts, we need to read in all the files, collate them, and then 
just grab the 'right now' and the 6h to compare sw/lw radiation.

```{r}
if (file.exists("data/herbie_extraction/compiled_forecast/GEFS_0p25_2023_3h_raw.feather")) {
  forecast_2023 <- read_feather("data/herbie_extraction/compiled_forecast/GEFS_0p25_2023_3h_raw.feather")
} else {
  forecast_files <- list.files("data/herbie_extraction/GEFS_2023/", full.names = T)
  forecast_2023 <- map(forecast_files, read_csv) %>% bind_rows()
  write_feather(forecast_2023, "data/herbie_extraction/compiled_forecast/GEFS_0p25_2023_3h_raw.feather")
}

```

Now, to SI:
- t2m is in deg K, needs to go to C
- u10,  v10 need to be transformed into wind speed

```{r}
forecast_2023_si <- forecast_2023 %>% 
  mutate(temp_degC = (t2m - 273.15),
         tmin = tmin-273.15,
         tmax = tmax-273.15,
         wind_mps = sqrt(u10^2 + v10^2), #sqrt(mer^2 + zon^2)
         solar_rad_wpm2 = sdswrf,
         rh = r2) %>%  
  select(-c(t2m, u10, v10, sdswrf, r2)) 

```

And drop a few more columns that we don't need to track:

```{r}
forecast_2023_si <- forecast_2023_si %>% 
  select(-c(heightAboveGround, gribfile_projection, surface)) %>% 
  # filter out a few stray na rows and a few oopsie 7d timesteps
  filter(!is.na(time), !grepl('7', step))

```

# Creating a timeseries of the forecast...

Here, we'll make the data 'look' like the met data we use in the workflow, then
aggregate as we do in the observed met. 

```{r}
rh_wind <-forecast_2023_si %>% 
  select(number, date = time, step, wind_mps, rh) %>% 
  mutate(days = str_sub(step, 1, 1),
         date = date(date),
         valid_date = date(date + days(days))) %>% 
  select(-c(step, days)) %>% 
  summarize(rh_min = min(rh, na.rm = T),
            rh_max = max(rh, na.rm = T),
            rh_ave = mean(rh, na.rm = T),
            wind_min = min(wind_mps, na.rm = T),
            wind_max = max(wind_mps, na.rm = T),
            wind_ave = ave(wind_mps, na.rm = T),
            .by = c('date', 'valid_date', 'number')) %>% 
  slice(1, .by = c('date', 'valid_date', 'number'))

air_temp <- forecast_2023_si %>% 
  select(number, date = time, step, temp_degC) %>% 
  mutate(days = str_sub(step, 1, 1),
         date = date(date),
         valid_date = date(date + days(days))) %>% 
  select(-c(step, days)) %>% 
  summarise(air_temp_min_3h = min(temp_degC, na.rm = T),
            air_temp_mean_3h = mean(temp_degC, na.rm = T),
            air_temp_max_3h = max(temp_degC, na.rm = T),
            .by = c('date', 'valid_date', 'number'))

temp_min_max <- forecast_2023_si %>% 
  select(number, date = time, step, tmax, tmin) %>% 
  mutate(days = str_sub(step, 1, 1),
         date = date(date),
         valid_date = date(date + days(days))) %>% 
  select(-c(step, days)) %>% 
  pivot_longer(cols = c('tmax', 'tmin'),
               names_to = 'parameter',
               values_to = 'value') %>% 
  summarise(air_temp_min_day = min(value, na.rm = T),
            air_temp_max_day = max(value, na.rm = T),
            .by = c('date', 'valid_date', 'number'))

sol_rad <- forecast_2023_si %>% 
  select(number, date = time, step, solar_rad_wpm2) %>% 
  mutate(days = str_sub(step, 1, 1),
         hour = str_sub(step, 8,9),
         date = date(date),
         valid_date = date(date + days(days))) %>% 
  filter(!hour %in% c('00', '03', '06')) %>% 
  select(-c(step, days, hour)) %>% 
  summarise(sol_rad_min_3h = min(solar_rad_wpm2, na.rm = T),
            sol_rad_max_3h = max(solar_rad_wpm2, na.rm = T),
            sol_rad_sum_3h = sum(solar_rad_wpm2, na.rm = T),
            .by = c('date', 'valid_date', 'number'))

timeseries <- reduce(list(air_temp, temp_min_max, sol_rad, rh_wind),
                     full_join)

write_csv(timeseries, "data/herbie_extraction/compiled_forecast/GEFS_0p25_2023_3h_summary.csv")
```

# Trasform and standardize

Transform the data:

```{r}
data_handle <- read_csv("data/NN_train_val_test/SMR_forecast/met_summary/data_handle_tracking_summary_t2023_v2024-11-29.csv")
params <- timeseries %>% 
  select(-c(date, valid_date, number)) %>% 
  names(.)
data_fore_handle <- data_handle %>% 
  filter(column %in% params) %>% 
  select(-`...1`)

forecast_v <- timeseries %>% 
  pivot_longer(c(air_temp_min_3h:wind_ave),
               names_to = "column",
               values_to = "value") %>% 
  left_join(., data_fore_handle)

forecast_trans <- forecast_v %>% 
  filter(!is.na(value)) %>% 
  mutate(value = case_when(is.na(shift) & transform == "square" ~ value**2,
                           is.na(shift) & transform == "square-root" ~ sqrt(value),
                           is.na(shift) & transform == "log" ~ log(value),
                           !is.na(shift) & transform == "square" ~ (value+shift)**2,
                           !is.na(shift) & transform == "square-root" ~ sqrt(value+shift),
                           !is.na(shift) & transform == "log" ~ log(value+shift),
                           .default = value)) %>% 
  select(-c(shift, transform)) %>% 
  pivot_wider(names_from = "column",
              values_from = "value") 

```

Standardize the data using the t2023 mean/std info:

```{r}
mean_std_t2023 <- read_csv("data/NN_train_val_test/SMR_forecast/met_summary/mean_std_train_val_summary_t2023_v2024-11-29.csv") 
names(mean_std_t2023) <- c("parameter", "mean", "std")

mean_std_t2023 <- mean_std_t2023 %>% 
  filter(parameter %in% unique(forecast_v$column))

standardize_cols <- function(data, parameter, mean, std) {
  param <- data[parameter]
  data[parameter] <- (param - mean) / std
  data %>% 
    select(date, valid_date, number, parameter)
}

forecast_s <- pmap(list(list(forecast_trans), 
                        mean_std_t2023$parameter,
                        mean_std_t2023$mean,
                        mean_std_t2023$std),
                   standardize_cols) %>% 
  reduce(., full_join, by = c("date", "valid_date", "number"))

```

And save the file for forecasting:

```{r}
write_csv(forecast_s, "data/NN_train_val_test/SMR_forecast/met_summary/t2023_forecast_std_v2024-12-01.csv")
```

