---
title: "01-Collate Forecast Data"
author: "ROSSyndicate"
date: "2024-11-22"
output: html_document
---

```{r}
library(tidyverse)
```

# Purpose

This markdown document collates 3h GEFS 0.25 data from our Herbie download.

For our forecasts, we need to read in all the files, collate them, and then 
just grab the 'right now' and the 6h to compare sw/lw radiation.

```{r}
forecast_files <- list.files("data/herbie_extraction/GEFS_2022/", full.names = T)

forecast_2022 <- map(forecast_files, read_csv) %>% bind_rows()

control_files <- list.files("data/herbie_extraction/debias/", full.names = T)

control_data <- map(control_files, read_csv) %>% bind_rows()

```

Now, to SI:
- t2m is in deg K, needs to go to C
- u10,  v10 need to be transformed into wind speed

```{r}
forecast_2022_si <- forecast_2022 %>% 
  mutate(temp_degC = (t2m - 273.15),
         wind_mps = sqrt(u10^2 + v10^2), #sqrt(mer^2 + zon^2)
         solar_rad_wpm2 = sdswrf,
         rh = r2) %>%  
  select(-c(t2m, u10, v10, sdswrf, r2)) 

control_si <- control_data %>% 
    mutate(temp_degC = (t2m - 273.15),
         wind_mps = sqrt(u10^2 + v10^2), #sqrt(mer^2 + zon^2)
         solar_rad_wpm2 = sdswrf,
         rh = r2) %>%  
  select(-c(t2m, u10, v10, sdswrf, r2)) 

```

And drop a few more columns that we don't need to track:

```{r}
forecast_2022_si <- forecast_2022_si %>% 
  select(-c(heightAboveGround, gribfile_projection, surface, `2022-06-01`)) %>% 
  filter(!is.na(time))

control_si <- control_si %>% 
  select(-c(heightAboveGround, gribfile_projection, surface, `2020-06-01`)) %>% 
  filter(!is.na(time))

```

# Creating a timeseries of the forecast...

temp, rh, wind is the forecast at valid time
tmin, tmax is over the 3h period since previous forecast
solar rad is the AVERAGE over 3h since previous forecast

## at forecast

for temp, rh, wind, we can just make some assumptions that there is a linear 
relationship between the previous and the valid. Let's just grab one example and
see if we can do this either linearly or using spline 

```{r}
int_test <- forecast_2022_si %>% 
  filter(time == ymd_hms("2022-06-02 06:00:00"), number == 0)

ggplot(int_test, aes(x = valid_time, y = temp_degC)) + 
  geom_point() +
  geom_line() +
  theme_bw() 
```


