---
title: "Forecast data integration with observed"
author: "ROSSyndicate"
date: "2024-10-30"
output: html_document
---

# Purpose

This markdown document is intended to determine how to handle the forecast
data obtained from NOAA GEFS using Herbie to integrate as part of the data
assimilation pipeline for operational forecasting of temperature in SMR using
the operational model described in Assignment 2.

## Load observed data and forecasted data

First, the observed met data - this is loaded from the NASA-NW repository
workflow.

```{r}
library(tidyverse)
library(feather)

w_SMR_today_met <- read_feather("~/Desktop/w_SMR_today_met.feather")
g_SMR_met_QAQC <- read_feather("~/Desktop/g_SMR_met_QAQC.feather")

```

For our forecasts, we need to read in all the files, collate them, and then 
just grab the 'right now' and the 6h to compare sw/lw radiation.

```{r}
all_forecast_files <- list.files("data/herbie_extraction/archive/", full.names = T)

forecast_files_2022 <- all_forecast_files[grepl("2022", all_forecast_files)]

forecast_2022 <- map(forecast_files_2022, read_csv) %>% bind_rows()

write_feather(forecast_2022, "data/herbie_extraction/compiled_forecast/GEFS_0p25_2022_raw.feather")

forecast_files_2023 <- all_forecast_files[grepl("2023", all_forecast_files)]

forecast_2023 <- map(forecast_files_2023, read_csv) %>% bind_rows()

write_feather(forecast_2023, "data/herbie_extraction/compiled_forecast/GEFS_0p25_2023_raw.feather")
```

We can drop a few columns here that we don't really care about

```{r}
forecast_control <- full_join(forecast_2022, forecast_2023) %>% 
  filter(number == 0) %>% 
  select(time, step, valid_time, t2m, u10, v10, sdswrf)
```

Now, to SI:
- t2m is in deg K, needs to go to C
- u10,  v10 need to be transformed into wind speed

```{r}
forecast_control <- forecast_control %>% 
  mutate(temp_degC = (t2m - 273.15),
         wind_mps = sqrt(u10^2 + v10^2), #sqrt(mer^2 + zon^2)
         solar_rad_wpm2 = sdswrf) %>% 
  select(-c(t2m, u10, v10, sdswrf)) 
```

## Look for biases in 6h (11p) data

```{r}
midnight <- g_SMR_met_QAQC %>% 
  filter(between(hour(datetime), 23, 24)) %>% 
        mutate(date = as_date(datetime) + days(1)) %>% # need to account for UTC around midnight 
        summarize(value = mean(value), .by = c(date, parameter)) %>% 
        filter(parameter %in% c("Air_Temperature_Avg", # t2m
                                "Wind_Speed_Avg")) %>% # gust
        pivot_wider(names_from = parameter,
                    values_from = value) %>% 
        rename(mid_air_temp = Air_Temperature_Avg,
               mid_ave_wind = Wind_Speed_Avg) #note, no solar rad at 0z
```

grab the control forecast

```{r}
control <- forecast_control %>% 
  filter(time == valid_time) %>% 
  mutate(date = date(time))
```

collate the control forecast with observed 

```{r}
bias_comp <- left_join(control, midnight)
```

And plot

```{r}
ggplot(bias_comp, aes(x = mid_air_temp, y = temp_degC)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0)

# lm debias
debias_lm = lm(bias_comp$temp_degC ~ bias_comp$mid_air_temp)
summary(debias_lm)

bias_comp$temp_debias_lm = (bias_comp$temp_degC + 0.97)/0.861

ggplot(bias_comp, aes(x = mid_air_temp, y = temp_debias_lm)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0)

```



```{r}
ggplot(bias_comp, aes(x = mid_ave_wind, y = wind_mps)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0)

# lm debias
wind_debias_lm = lm(bias_comp$wind_mps ~ bias_comp$mid_ave_wind)
summary(wind_debias_lm)

bias_comp$wind_debias_lm = (bias_comp$wind_mps - 1.5)/0.27

ggplot(bias_comp, aes(x = mid_ave_wind, y = wind_debias_lm)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0)

# debias by eye
bias_comp$wind_debias_eye = bias_comp$wind_mps/2

ggplot(bias_comp, aes(x = mid_ave_wind, y = wind_debias_eye)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0)

```


And now we'll look at the control at 12h for solar radiation. This isn't perfect,
but it gives us an idea of approximate bias

```{r}
control_rad <- forecast_control %>% 
  filter(step == "0 days 12:00:00") %>% 
  mutate(date = date(time))

bias_comp_rad <- left_join(control_rad, w_SMR_today_met)

ggplot(bias_comp_rad, aes(x = noon_solar_rad, y = solar_rad_wpm2)) +
  geom_point()

# lm debias
solar_rad_debias_lm = lm(bias_comp_rad$solar_rad_wpm2 ~ bias_comp_rad$noon_solar_rad)
summary(solar_rad_debias_lm)

bias_comp_rad$solar_rad_debias_lm = (bias_comp_rad$solar_rad_wpm2 - 179.7)/0.368

ggplot(bias_comp_rad, aes(x = noon_solar_rad, y = solar_rad_debias_lm)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0)

# by eye debias, given coordinates (500, 250) and (1000, 600), slope is 0.7 and intercept is -100

bias_comp_rad$solar_rad_debias_eye =  (bias_comp_rad$solar_rad_wpm2 + 100)/0.7

ggplot(bias_comp_rad, aes(x = noon_solar_rad, y = solar_rad_debias_eye)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0)

```

Great! Let's move all this to a different script and work on creating a function to integrate observed and forecasted data.

```{r}
debiased_herbie <- full_join(forecast_2022, forecast_2023) %>% 
  mutate(temp_degC = ((t2m - 273.15) + 0.97)/0.861,
         wind_mps = sqrt(u10^2 + v10^2)/2, #sqrt(mer^2 + zon^2)
         solar_rad_wpm2 = (sdswrf + 100)/0.7) %>% 
  select(number, time, step, valid_time, temp_degC, wind_mps, solar_rad_wpm2)

write_feather(debiased_herbie, "data/herbie_extraction/compiled_forecast/simple_debias_2022_2023.feather")
```

# format and standardize the forecast

```{r}
forecast_met <- debiased_herbie %>% 
  filter(between(date(time), ymd("2022-05-01"), ymd("2022-11-01"))) %>% 
  filter(step != "0 days 00:00:00") %>% 
  mutate(forecast_date = date(time),
         date = date(valid_time)) %>% 
  select(-c(time, step, valid_time)) %>% 
  rename(noon_air_temp = temp_degC, 
         noon_ave_wind = wind_mps, 
         noon_solar_rad = solar_rad_wpm2)
```

Standardize the forecast data for 2022 - this way, no transformation needs to 
be done within the forecast function itself, since the inputs and outputs are 
standardized. 

Transform the data:

```{r}
forecast_less <- forecast_met %>% 
  mutate(noon_air_temp = noon_air_temp^2,
         noon_ave_wind = log(noon_ave_wind),
         noon_solar_rad = noon_solar_rad^2)
```

Standardize the data using the t2022 mean/std info:

```{r}
mean_std_t2022 <- read_csv("data/NN_train_val_test/SMR_forecast/mean_std_train_val_t2022_v2024-10-28.csv") 
names(mean_std_t2022) <- c("parameter", "mean", "std")

mean_std_t2022 <- mean_std_t2022 %>% 
  filter(grepl("noon_", parameter), !grepl("_m", parameter))

standardize_cols <- function(data, parameter, mean, std) {
  param <- data[parameter]
  data[parameter] <- (param - mean) / std
  data %>% 
    select(date, forecast_date, number, parameter)
}

forecast_s <- pmap(list(list(forecast_less), 
                        mean_std_t2022$parameter,
                        mean_std_t2022$mean,
                        mean_std_t2022$std),
                   standardize_cols) %>% 
  reduce(., full_join, by = c("date", "forecast_date", "number"))
```

And save the file for forecasting:

```{r}
write_csv(forecast_s, "data/NN_train_val_test/SMR_forecast/t2022_forecast_std_v2024-10-28.csv")
```

