---
title: "Access NOAA GEFS Forecasts"
author: "ROSSyndicate"
date: "2024-09-19"
output: html_document
---

# Purpose

This script walks through acquisition of archived NOAA GEFS forecasts for the
Three Lakes System. This takes about 7m per date at 0.5deg resolution and closer 
to 30m for 0.25 resolution. After playing around, it is more efficient to use
Herbie in a Python environment (19m per date for 0.25 deg resolution), so just
archiving this as a how-to and/or intro, but not as 'functional' code per-se.


``` {r}
library(tidyverse)
library(magick)
library(aws.s3)
library(tmap)
```

# Intro to AWS

Check for the bucket

```{r}
bucket_exists(
  bucket = "s3://noaa-gefs-pds/",
  region = "us-east-1"
  )
```

Look at the contents

```{r}
get_bucket_df(
  bucket = "s3://noaa-gefs-pds/",
  region = "us-east-1",
  prefix = "gefs.20201001/00/atmos/pgrb2sp25/",
  max = 100
) %>% 
  as_tibble()
```

Can we download something?

0.25 resolution starts ~ Oct 2020
https://www.nco.ncep.noaa.gov/pmb/products/gens/gep01.t00z.pgrb2s.0p25.f000.shtml

```{r}
# if object exists, save it and return file path
save_object(
  object = "gefs.20201001/00/atmos/pgrb2sp25/gep01.t00z.pgrb2s.0p25.f000",
  bucket = "s3://noaa-gefs-pds/",
  region = "us-east-1",
  file = "test.grb2"
)
```

Load it?

```{r}
test_gefs <- read_stars("test.grb2")
```

Plot it?

```{r}
smaller_gefs <- test_gefs %>% 
  filter(x < -105) %>% 
  filter(x > -106) %>% 
  filter(y > 40) %>% 
  filter(y < 41)

plot(smaller_gefs)

smr <- st_point(c(-105.85, 40.22)) %>% 
  st_sfc(., crs = "EPSG:4326")

tmap_mode("view")
tm_shape(smr) +
  tm_dots()

smr_extr <- st_extract(smaller_gefs, st_transform(smr, crs = st_crs(smaller_gefs))) %>% 
  st_as_sf() %>% 
  st_drop_geometry()

colnames(smr_extr) <- c("GUST", "PRES", "HGT", "TSOIL", "SOILW", "WEASD", "SNOD",
                        "ICETK", "TMP", "DPT", "RH", "TMAX", "TMIN", "UGRD", "VGRD",
                        "CAPE", "CIN", "PWAT", "HLCY", "CAPE2", "CIN2", "PRMSL")
```

The 0.25km has about everything we need except SW/LW Radiation. So let's also 
grab those data from the 0.5 data (we'll grab the rest of the data from those
too, just to compare to see how worth it is to grab 0.25deg and 0.5deg data)

```{r}
get_bucket_df(
  bucket = "s3://noaa-gefs-pds/",
  region = "us-east-1",
  prefix = "gefs.20201001/00/atmos/pgrb2ap5/",
  max = 100
) %>% 
  as_tibble()
```

```{r}
# if object exists, save it and return file path - 1800 utc is ~ noon. 
save_object(
  object = "gefs.20201001/06/atmos/pgrb2ap5/gep01.t06z.pgrb2a.0p50.f018",
  bucket = "s3://noaa-gefs-pds/",
  region = "us-east-1",
  file = "test.grb2"
)

test_gefs <- read_stars("test.grb2")

smr_extr <- st_extract(test_gefs, st_transform(smr, crs = st_crs(test_gefs))) %>% 
  st_as_sf() %>% 
  st_drop_geometry() 

names(smr_extr)[c(60:68, 78, 79)] <- c("WEASD", "SNOD", "ICETK", "TMP", "RH", "TMAX", "TMIN", "UGRD", "VGRD", "DSWRF", "DLWRF")

fewer_extr <- smr_extr %>% select("WEASD", "SNOD", "ICETK", "TMP", "RH", "TMAX", "TMIN", "UGRD", "VGRD", "DSWRF", "DLWRF")

```

We want to access 7-day forecasts for noon every 24h at midnight local time (06 UTC).

Function will change date. Folder will always be 06 (midnight local) and then iterate 12, 36, 60, 84, 108, 132, 156 grabbing data. Need to grab each control + perturbation.

