---
title: "Feature Reduction (version 2)"
author: "ROSSyndicate"
date: "2024-05-23"
output: html_document
---

```{r}
library(tidyverse)
library(GGally)

upstream_version = '2024-05-21'
reduced_version = '2024-06-03'
```

# Purpose

After trying the aggressive feature reduction, I'll try something in between. The
arNN of the first pass of reduced features performed more poorly than yesterday-is-today,
and it might be because we've dropped parameters that are important (even if not
as important as those listed in the SHAP). Also, I noticed a few redundant features
that we should eliminate from the dataset.

## Feature Reduction

We'll keep the inflow data this time, but reduce the redundant features in the
same way.

```{r}
t2022 <- read_csv('data/NN_train_val_test/SMR_autoNN_daily/trainval_t2022_v2024-05-09.csv')

t2022_reduce <- t2022
```

Let's look at some correlation plots to determine if we should toss any
redundant features, first by data group, since that's likely where we'll see
some redundancy.

### Yesterday's water temperature

```{r}
col_names = names(t2022_reduce)

yesterday <- t2022_reduce %>% 
  select(date, all_of(col_names[grepl("y_", col_names)])) 

ggpairs(yesterday, columns = 2:ncol(yesterday))
```

Max 1 and max 0-5 are the same (which makes sense)

```{r}
yesterday_reduced <- yesterday %>% 
  select(-y_max_1m_temp_degC)

ggpairs(yesterday_reduced, columns = 2:ncol(yesterday_reduced))
```

We could probably reduce this further, so let's do that by dropping all the 1m
summaries

```{r}
yesterday_reduced <- yesterday_reduced %>% 
  select(-c(y_min_1m_temp_degC, y_mean_1m_temp_degC))

ggpairs(yesterday_reduced, columns = 2:ncol(yesterday_reduced))
```

### Air Temperature

```{r}
temp <- t2022_reduce %>% 
  select(date, all_of(col_names[grepl('temp', col_names)])) %>% 
  select(-all_of(col_names[grepl('1m|5m', col_names)])) 

ggpairs(temp, columns = 2:ncol(temp))
```

These are all \< 0.9 correlation, so let's leave it for now.

### Solar Radiation

```{r}
solrad <- t2022_reduce %>% 
  select(date, all_of(col_names[grepl('sol', col_names)]))

ggpairs(solrad, columns = 2:ncol(solrad))
```

These look good too.

### Pump

```{r}
pump <- t2022_reduce %>% 
  select(date, all_of(col_names[grepl('pump', col_names)]))

ggpairs(pump, columns = 2:ncol(pump))
```

Oh, right - there is a linear relationship between the average and the sum over
a designated time period. We can toss the mean in favor of the sum.

```{r}
pump_reduced <- pump %>% 
  select(-c(mean_pump_q_p2, mean_pump_q_p7))

ggpairs(pump_reduced, columns = 2:ncol(pump_reduced))
```

### Wind Speed

```{r}
wind <- t2022_reduce %>% 
  select(date, all_of(col_names[grepl('wind', col_names)]))

ggpairs(wind, columns = 2:ncol(wind))
```

Min wind 3/5, Max wind 3/5, Min 10/5 - we can drop max/min 5 and solve these
redundancies.

```{r}
wind_reduced <- wind %>% 
  select(-c(max_wind_mps_5, min_wind_mps_5))

ggpairs(wind_reduced, columns = 2:ncol(wind_reduced))
```

### Precip

```{r}
precip <- t2022_reduce %>% 
  select(date, all_of(col_names[grepl('precip', col_names)]))

ggpairs(precip, columns = 2:ncol(precip))
```

Looks good. So dry.

### North Fork

```{r}
NF <- t2022_reduce %>% 
  select(date, all_of(col_names[grepl('NF', col_names)]))

ggpairs(NF, columns = 2:ncol(NF))
```

Similar issue here where there are sum and averages that are 1:1, also dropping 
the other p2 parameters since they seem to have high correlation with minus 1 
and minus 2 day values.

```{r}
NF_reduced <- NF %>% 
  select(-c(sum_NF_q_p2, sum_NF_q_p7, max_NF_q_p2, mean_NF_q_p2))

ggpairs(NF_reduced, columns = 2:ncol(NF_reduced))

```

### Chipmunk Lane

```{r}
chip <- t2022_reduce %>% 
  select(date, all_of(col_names[grepl('chip', col_names)]))

ggpairs(chip, columns = 2:ncol(chip))
```

Remove the sum/average dupes

```{r}
chip_reduced <- chip %>% 
  select(-c(sum_chip_q_p2, sum_chip_q_p7, max_chip_q_p2, mean_chip_q_p2))

ggpairs(chip_reduced, columns = 2:ncol(chip_reduced))
```


## Collate and export by year

```{r}
# collate data
t2022_reduced <- reduce(list(yesterday_reduced, pump_reduced, temp, wind_reduced, solrad, precip, NF_reduced, chip_reduced),
                        full_join)

reduce_names = names(t2022_reduced)

# add back in the labels
reduce_names = append(reduce_names, c("mean_1m_temp_degC", "mean_0_5m_temp_degC"))

years = seq(2014, 2021)

og = 'data/NN_train_val_test/SMR_autoNN_daily'
fp = 'data/NN_train_val_test/SMR_autoNN_reduce_2'
map(.x = years,
    .f = ~ {
      validation <- read_csv(file.path(og,
                                       paste0('validation_t2022_',
                                              .x,
                                              '_v',
                                              upstream_version, 
                                              '.csv'))) %>% 
        select(all_of(reduce_names))
      write_csv(validation, 
                file.path(fp, 
                          paste0("validation_t2022_",
                                 .x,
                                 "_reduced_v",
                                 reduced_version,
                                 ".csv")))
      training <- read_csv(file.path(og,
                                       paste0('training_t2022_',
                                              .x,
                                              '_v', 
                                              upstream_version,
                                              '.csv'))) %>% 
        select(all_of(reduce_names))
      write_csv(training, 
                file.path(fp, 
                          paste0("training_t2022_",
                                 .x,
                                 "_reduced_v",
                                 reduced_version,
                                 ".csv")))
    })
```

And also subset the test set

```{r}
test <- read_csv(file.path(og, paste0('t2022_standardized_v',
                                      upstream_version, 
                                      '.csv')))

test_reduced <- test %>% 
  select(reduce_names)

write_csv(test_reduced, file.path(fp, paste0('t2022_reduced_standardized_v', reduced_version, '.csv')))
```
